State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2020-09-01 analysis/paper/paper.Rmd
Better described how the trap rate was estimated and fixed a typo in the caption of figure showing relative bias in simulations.

A trap rate estimate was derived using mark-recapture methods and PIT tag observations of both Chinook Salmon and steelhead at Lower Granite Dam adult detection sites. The "mark" group included all tags detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system. Using a mark-recapture model with differing capture probabilities, we estimated the trap rate on a weekly basis. Those estimates, with associated uncertainties, were then incorporated into the model as informed priors, while the model estimated the true trap rate based on all the data, including trap and window counts. We want to estimate the "true" trapping rate as the set trapping rate, the recorded time that the trap is open to trap adults, does not always reflect the true proportion of fish that are captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others. Therefore, we use the mark-recapture estimate of the trap rate to estimate a true trapping rate.

@@

We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage, and fallback/re-ascension rates using a state-space modeling approach [@royle2008hierarchical] implemented in the STADEM package for the R statistical software [@R-Core-Team2020]. We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error. In the case of the trap, for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate. STADEM further accounted for the proportion of fish that ascended the ladder while the counting window was closed (i.e., night), as well as for fish potentially double-counted (or more) after falling back below the dam and later re-ascending the fish ladder. Finally, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig)). Additional model details can be found in [Appendix A](#append1). The STADEM package is available from the primary author at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM), and requires the use of the JAGS software [@R-rjags] for Bayesian inference.

@@

```{r rel-bias-lgd, fig.cap = "'Boxplots show relative bias of STADEM estimates for wild escapement across various scenarios. The boxes contain 50% of the simulations, whiskers contain 95% of the simulations, and points are outliers."}
hdi_box = function(x) {
  r = c(hdi(x, 0.95), hdi(x, 0.5), median(x))
  # r = quantile(x, probs = c(0.025, 0.975, 0.25, 0.75, 0.5))
  names(r) = c('ymin', 'ymax', 'lower', 'upper', 'middle')
  r
}

>>>>

A trap rate estimate was derived using mark-recapture methods and PIT tag observations of both Chinook Salmon and steelhead at Lower Granite Dam adult detection sites. The "mark" group included all tags detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system. Using a mark-recapture model with differing capture probabilities, we estimated the trap rate on a weekly basis. Those estimates, with associated uncertainties, were then incorporated into the model as informed priors, while the model estimated the true trap rate based on all the data, including trap and window counts. We want to estimate the "true" trapping rate because the set trapping rate (i.e. the recorded time that the trap is open to trap adults) does not always reflect the true proportion of fish that are captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others. Therefore, we use the mark-recapture estimate of the trap rate to estimate a true trapping rate.

<!-- Fix the first two sentences below to make it clear the rates are estimated as a binomial with a rate that has a random walk in logit space -->

@@

We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage and fallback/re-ascension rates using a state-space modeling approach [@royle2008hierarchical] implemented in the STADEM package for the R statistical software [@R-Core-Team2020]. We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error. In the case of the trap, for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate. STADEM further accounted for the proportion of fish that ascended the ladder while the counting window was closed (i.e., night), as well as for fish potentially double-counted (or more) after falling back below the dam and later re-ascending the fish ladder. Finally, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig)). Additional model details can be found in [Appendix A](#append1). The STADEM package is available from the primary author at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM), and requires the use of the JAGS software [@R-rjags] for Bayesian inference.

@@

```{r rel-bias-lgd, fig.cap = "Boxplots show relative bias of STADEM estimates for wild escapement across various scenarios. The boxes contain 50% of the simulations, whiskers contain 95% of the simulations, and points are outliers."}
hdi_box = function(x) {
  r = c(hdi(x, 0.95), hdi(x, 0.5), median(x))
  # r = quantile(x, probs = c(0.025, 0.975, 0.25, 0.75, 0.5))
  names(r) = c('ymin', 'ymax', 'lower', 'upper', 'middle')
  r
}
